### 2014-10-29 contract-language.tex
Edits to Secs 4 and 5.

\subsection{Capturing Store Semantics}
\label{sec:store_sem}

An important aspect of our contract classification system is that the store
semantics is also captured using the same contract language used to describe
application-level consistency. In this regard, similar
to~\cite{Burckhardt2014}, we can rigorously define a wide variety of store
semantics including those that combine any subset of session and causality
guarantees, and multiple consistency levels. For example, a store that offers
strong consistency is captured by the contract:

\vspace{-1em}
\begin{smathpar}
\scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
\end{smathpar}

\noindent Similarly, a store that offers per-object causal consistency is captured by the
contract:

\vspace{-1em}
\begin{smathpar}
\ccc = \forall a.~(\hboZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff}
\end{smathpar}

\noindent This ability to represent store semantics and application-level consistency in
the same language is vital to contract classification. Observe that out
contract language does not incorporate real (i.e., wall-clock) time. Hence, the
contract language cannot describe store semantics based on real time such as
recency or bounded-staleness guarantees offered by certain
stores~\cite{Pileus}.

\subsection{Contract Comparison}

Our goal is to classify contracts, and map the operation they describe to
the \emph{weakest} store-level consistency level that  nonetheless
satisfies the contract's constraints. To this end, we need a mechanism to
compare the ``strength'' of a contract. Let $\cv_{op}$ be a contract for a
particular operation $op$, and $\cv_{st}$ capture a particular store
consistency level. We would like to determine whether $op$ can be
\emph{safely discharged} at the store consistency level $\cv_{st}$ such that
the resulting execution does not violate $\cv_{op}$.

Since our contracts represent axiomatic definition of program executions, let
$\Mod{\cv}$ be the set of all executions under which contract $\cv$ is
satisfied. If every execution $\E \in \Mod{\cv_{st}}$ is also a member of
$\Mod{\cv_{op}}$, then $op$ can be safely discharged under the store
consistency level $\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq
\Mod{\cv_{op}}$. This is the model-theoretic consequence relation, written as
$\cv_{st} \models_m \cv_{op}$.

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}.  Since first-order logic is sound and complete
~\cite{completeness}, $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st}
\Rightarrow \cv_{op}$.  Due to the decidability of our contract language, this
implication check is automatically discharged with the help of a theorem
prover.

\subsection{Contract Classification}

While the store semantics offered by commercial and research data stores
vary widely, for our purposes, we identify three particular consistency
levels -- eventual, causal, and strong, commonly offered by many distributed
stores with tunable consistency, with increasing overhead in terms of
latency and availability. Indeed, the techniques presented here can be
extended to other consistency stratifications. We assign each
application-level contract into one of these following classes:

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
  be satisfied as long as the client can reach at least one replica. For
  example, \cf{deposit} is an eventually consistent operation; its semantics
  does not require its action to manifest on all replicas before other
  operations in its session are allowed to proceed. While eventually
  consistent data store typically offer \emph{basic} eventual consistency
  with all possible anomalies, we assume that our store provides stronger
  semantics that remain highly-available~\cite{BailisHAT,COPS}; the store
  always exposes a causal cut of the updates. This semantics can be formally
  captured in terms of the following contract definition:

\begin{smathpar}
\ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}
\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
	contracts are required to see a causally consistent snapshot of the object
	state, including the actions performed on the same session.  The latter
	requirement entails that if two operations $o_1$ and $o_2$ from the same
	session are applied to two different replicas $r_1$ and $r_2$, the second
	operation cannot be discharged until the effect of $o_1$ is merged with $o_2$
	in both $r_1$ and $r_2$. The \cf{getBalance} operation requires causal
	consistency, as it requires the operations from the same session to be
	visible, which cannot be guaranteed under eventual consistency. We assume
	that causality is only tracked through operations on the same object; two
	operations in the same session but on different objects are considered
	causally unrelated under this definition. Stores typically avoid tracking
	causality across objects to mitigate overheads when causality tracking is
	unnecessary. The corresponding store semantics is captured by the contract
	$\ccc$ presented in Section~\ref{sec:store_sem}.

\item \textbf{Strong Consistency}: Strongly consistent operations may block
  indefinitely under network partitions. An example is the total-order
  contract on \cf{withdraw} operation. The corresponding store semantics is
  captured by the contract $\scc$ presented in Section~\ref{sec:store_sem}.

@@

\noindent Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves
totally ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$.
This concurs with the intuition that any contract satisfiable under $\ecc$ or
$\ccc$ is satisfiable under $\scc$, and any contract that is satisfiable under
$\ecc$ is satisfiable under $\ccc$. Nonetheless, we determine the class of a
contract based on the \emph{weakest} guarantee (among $\ecc$, $\ccc$, and
$\scc$) required to satisfy the contract.

@@

The classification scheme is presented formally in Figure~\ref{sem:classify}.
Along with three straightforward rules that classify contracts into consistency
classes, the classification scheme also presents a rule that judge
well-formedness of a contract. A contract is well-formed if and only if it is
satisfiable under $\scc$ - the strongest possible consistency guarantee any
store can provide. The $\le$ relation used by the rules is defined formally
below:

\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-1em}

The $\Delta$ referred in the above defintion captures assumptions about the
nature of primitive relations, such as $\Rvis$ and $\Rso$ are irreflexive, and
the happens-before relation $\hbZ$ is acyclic, preventing thin-air reads.
\KC{Ideally, we need to list all the axioms somewhere.}

>>>>

% \subsection{Capturing Store Semantics}
% \label{sec:store_sem}

% An important aspect of our contract classification system is that the store
% semantics is also captured using the same contract language used to describe
% application-level consistency. In this regard, similar
% to~\cite{Burckhardt2014}, we can rigorously define a wide variety of store
% semantics including those that combine any subset of session and causality
% guarantees, and multiple consistency levels. For example, a store that offers
% strong consistency is captured by the contract:

% \vspace{-1em}
% \begin{smathpar}
% \scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
% \end{smathpar}

% \noindent Similarly, a store that offers per-object causal consistency is captured by the
% contract:

% \vspace{-1em}
% \begin{smathpar}
% \ccc = \forall a.~(\hboZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff}
% \end{smathpar}

% \noindent This ability to represent store semantics and application-level consistency in
% the same language is vital to contract classification.

% \subsection{Contract Comparison}

% Our goal is to classify contracts, and map the operation they describe to
% the \emph{weakest} store-level consistency level that  nonetheless
% satisfies the contract's constraints. To this end, we need a mechanism to
% compare the ``strength'' of a contract. Let $\cv_{op}$ be a contract for a
% particular operation $op$, and $\cv_{st}$ capture a particular store
% consistency level. We would like to determine whether $op$ can be
% \emph{safely discharged} at the store consistency level $\cv_{st}$ such that
% the resulting execution does not violate $\cv_{op}$.

% Since our contracts represent axiomatic definition of program executions, let
% $\Mod{\cv}$ be the set of all executions under which contract $\cv$ is
% satisfied. If every execution $\E \in \Mod{\cv_{st}}$ is also a member of
% $\Mod{\cv_{op}}$, then $op$ can be safely discharged under the store
% consistency level $\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq
% \Mod{\cv_{op}}$. This is the model-theoretic consequence relation, written as
% $\cv_{st} \models_m \cv_{op}$.

% Observe that our contract language (stripped of its syntactic sugar) is a
% carefully chosen subset of first-order logic that is known to be
% decidable~\cite{epr}.  Since first-order logic is sound and complete
% ~\cite{completeness}, $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st}
% \Rightarrow \cv_{op}$.  Due to the decidability of our contract language, this
% implication check is automatically discharged with the help of a theorem
% prover.

\subsection{Capturing Store Semantics}

An important aspect of our contract language is its ability to capture
store-level consistency guarantees, along with application-level
consistency requirements. Similar to~\cite{Burckhardt2014}, we can
rigorously define a wide variety of store semantics including those
that combine any subset of session and causality guarantees, and
multiple consistency levels.  However, for our purposes, we identify
three particular consistency levels -- eventual, causal, and strong,
commonly offered by many distributed stores with tunable consistency,
with increasing overhead in terms of latency and availability. For
each of the these three consistency levels, we capture the semantics
as a formula in our contract language, and informally describe the kind
of application-level consistency requirements that can met under the
consistency level:

% Indeed, the techniques presented here can be
% extended to other consistency stratifications. We assign each
% application-level contract into one of these following classes:

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
  be satisfied as long as the client can reach at least one replica. For
  example, \cf{deposit} is an eventually consistent operation; its semantics
  does not require its action to manifest on all replicas before other
  operations in its session are allowed to proceed. While eventually
  consistent data store typically offer \emph{basic} eventual consistency
  with all possible anomalies, we assume that our store provides stronger
  semantics that remain highly-available~\cite{BailisHAT,COPS}; the store
  always exposes a causal cut of the updates. This semantics can be formally
  captured in terms of the following contract definition:
  \vspace{-0.8em}
  \begin{smathpar}
  \ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
  \end{smathpar}
  \noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Causally consistent operations
  require to see a causally consistent snapshot of the object state,
  including the actions performed on the same session.  The latter
  requirement entails that if two operations $o_1$ and $o_2$ from the
  same session are applied to two different replicas $r_1$ and $r_2$,
  the second operation cannot be discharged until the effect of $o_1$ is
  merged with $o_2$ in both $r_1$ and $r_2$. The \cf{getBalance}
  operation requires causal consistency, as it requires the operations
  from the same session to be visible, which cannot be guaranteed under
  eventual consistency. We assume that causality is only tracked through
  operations on the same object; two operations in the same session but
  on different objects are considered causally unrelated under this
  definition. Stores typically avoid tracking causality across objects
  to mitigate overheads when causality tracking is unnecessary. The
  corresponding store semantics is captured by the contract $\ccc$
  defined below:
  \vspace{-0.8em}
  \begin{smathpar}
  \ccc = \forall a.~(\hboZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff}
  \end{smathpar}

\item \textbf{Strong Consistency}: Strongly consistent operations may block
  indefinitely under network partitions. An example is the total-order
  contract on \cf{withdraw} operation. The corresponding store semantics is
  captured by the $\scc$ contract definition:
  \vspace{-0.8em}
  \begin{smathpar}
  \scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
  \end{smathpar}

@@

Observe that out contract language does not incorporate real (i.e.,
wall-clock) time. Hence, the contract language cannot describe store
semantics based on real time such as recency or bounded-staleness
guarantees offered by certain stores~\cite{Pileus}.

@@

\subsection{Contract Comparison and Classification}

Our goal is to map application-level consistency constraints of
operations to appropriate store-level consistency guarantees capable
of satisfying constraints.  The ability to express both of them as
contracts in our contract language lets us compare and determine if
contract ($\cv_{op}$) of an operation ($op$) is weak enough to be
satisfied under a store consistency level identified by the contract
$\cv_{st}$. Towards this end, we define a binary \emph{weaker than}
relation for our contract language as following:
\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}

\vspace{-2em} \noindent The $\Delta$ referred in the above defintion captures
assumptions about the nature of primitive relations, such as $\Rvis$
and $\Rso$ are irreflexive, and the happens-before relation $\hbZ$ is
acyclic, preventing thin-air reads.

If the contract ($\cv_{op}$) of an operation ($op$) is \emph{weaker
than} a store contract ($\cv_{st}$), then constraints expressed by the
former are implied by guarantees provided by the later. Consequently,
$\cv_{op}$ is satisfiable under $\cv_{st}$. Observe that the contracts
$\scc$, $\ccc$ and $\ecc$ are themselves totally ordered with respect
to the $\le$ relation: $\ecc \le \ccc \le \scc$.  This concurs with
the intuition that any contract satisfiable under $\ecc$ or $\ccc$ is
satisfiable under $\scc$, and any contract that is satisfiable under
$\ecc$ is satisfiable under $\ccc$. However, we are interested in the
\emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$) required
to satisfy the contract. We define the corresponding consistency level
as the \emph{consistency class of the contract}. The classification
scheme is presented formally in Figure~\ref{sem:classify}.

Along with three straightforward rules that classify contracts into
consistency classes, the classification scheme also presents a rule
that judges well-formedness of a contract. A contract is well-formed
if and only if it is satisfiable under $\scc$ - the strongest possible
consistency guarantee any store can provide. Otherwise, it is
considered ill-formed, and rejected statically.

### 2014-10-29 core-lang.tex
Edits
Edits to Secs 4 and 5.

Fig.~\ref{sem:oper} summarizes the artifacts needed to state our
meta-theoretic result. We formalize an abstract model of a replicated
data store as a tuple $\E;\Sigma$, where $\E$ denotes the execution
state, and $\Sigma$, the session soup, is the set of concurrent client
sessions interacting with the store. The execution state ($\E$) of a
data store is modeled as a tuple $\Exec$ where $\EffSoup$ is a set of
effects during the execution, and $\visZ,\soZ \subseteq \EffSoup
\times \EffSoup$ are visibility and session order relations,
respectively, witnessed over generated effects. A session is modeled
simply as a sequence of replicated data type operations ($op$), tagged
with the consistency class ($\tau$) of their contracts (as determined
by the contract classification scheme). We abbreviate contract classes
($\tau$) as uppercase {\sf SC}, {\sf CC}, and {\sf EC} for brevity.
Their representative contracts ($\cv(\tau)$) are, as usual, written
$\scc$, $\ccc$, and $\ecc$, respectively.  Our contract language and
classification scheme is independent of store semantics. Accordingly,
we abstain from making any specific assumptions about semantics of the
store. However, we do assume the existence of a reduction relation of
form:
\begin{smathpar} 
  \auxred{} {(\E,\Sigma)} {\eff} {(\E',\Sigma')}
\end{smathpar}
\noindent The relation captures store's progress in execution (from
$\E$ to $\E'$) due to the application of an operation from one of the
sessions in $\Sigma$ to a replica, generating a new effect ($\eff$).

@@

\begin{theorem}[Soundness of Contract Enforcement]
\label{lem:core-preservation}
Let $\cv$ be the contract of a replicated data type operation $op$,
and let $\tau$ denote the consistency class of $\cv$ 
as determined by the contract classification scheme. Forall execution
states $\E$, $\E'$ such that 
$\auxred{} {(\E,\langle (op,\tau);\,\sigma \rangle \pll \Sigma)} {\eff} 
 {(\E',\langle \sigma \rangle \pll \Sigma)}$
if $\E' \models [\eff/\cureff]\,\cv(\tau)$, then $\E' \models [\eff/\cureff]\,\cv$ 
\end{theorem}

The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ guarantees in all executions, then contract
classification scheme can be used to satisfy any \name contract on
that data store. The proof of the theorem has been included in our
tech report~\cite{techrep}. Tech report also contains operational
semantics of our implementation of the data store, which concretizes
the reduction relation ($\xhookrightarrow{}$), along with the proof
that it correctly enforces all \name contracts.

>>>>

Fig.~\ref{sem:oper} summarizes the artifacts needed to state our
meta-theoretic result. We formalize an abstract model of a replicated
data store as a tuple $\E;\Sigma$, where $\E$ denotes the execution
state, and $\Sigma$, the session soup, is the set of concurrent client
sessions interacting with the store. The execution state ($\E$) of a
data store is modeled as a tuple $\Exec$ where $\EffSoup$, called the
\emph{effect soup}, is a set of effects generated during the
execution, and $\visZ,\soZ \subseteq \EffSoup \times \EffSoup$ are
visibility and session order relations, respectively, witnessed over
generated effects. A session is modeled simply as a sequence of
replicated data type operations ($op$), tagged with the consistency
class ($\tau$) of their contracts (as determined by the contract
classification scheme). Our contract language and classification
scheme is independent of store semantics. Accordingly, we abstain from
making any specific assumptions about semantics of the store. However,
we do assume the existence of a reduction relation of form:

\begin{smathpar}
  \auxred{} {\E;\langle (op,\tau);\sigma \rangle \pll \Sigma} {\eff} 
    {\E';\langle \sigma \rangle \pll \Sigma'}
\end{smathpar}

\noindent The relation captures store's progress in execution (from
$\E$ to $\E'$) due to the application of an operation ($op$) from one
of the sessions in $\Sigma$ to a replica, generating a new effect
($\eff$).

@@

\begin{theorem}[Soundness of Contract Enforcement]
\label{lem:core-preservation}
Let $\cv$ be the contract of a replicated data type operation $op$,
and let $\tau$ denote the consistency class of $\cv$
as determined by the contract classification scheme. Forall execution
states $\E$, $\E'$ such that
$\auxred{} {\E;\langle (op,\tau);\,\sigma \rangle \pll \Sigma} {\eff}
 {\E';\langle \sigma \rangle \pll \Sigma}$
if $\E' \models [\eff/\cureff]\,\cv(\tau)$, then $\E' \models [\eff/\cureff]\,\cv$
\end{theorem}

The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ guarantees in all executions, then contract
classification scheme can be used to satisfy any \name contract on
that data store. The proof of the theorem has been included in our
tech report~\cite{techrep}. Tech report also contains operational
semantics of our implementation of the data store, which concretizes
the reduction relation ($\xrightarrow{}$), along with the proof
that it correctly enforces all \name contracts.
